---
title: "plasma storage project - data analysis"
author: "Savannah Weaver"
date: "2023"
output: 
  rmdformats::html_clean:
    highlight: tango
    thumbnails: FALSE
    toc: TRUE
    toc_depth: 3
---


# Packages

```{r setup, include = TRUE}
if (!require("tidyverse")) install.packages("tidyverse")
library("tidyverse") # workflow and plots
if (!require("rmdformats")) install.packages("rmdformats")
library("rmdformats") # clean html R markdown format
if (!require("emmeans")) install.packages("emmeans")
library("emmeans") # marginal means
```



# Background and Goals

We will take our cleaned-up df from the wrangling Rmd, and make some figures and run some statistics to determine whether storage temperature and time affect the plasma osmolality of blood plasma samples. 





# Load Data

```{r}
dat <- read_rds("./data/osml_means_clean.RDS") %>% 
  mutate(Temp_Time = as.factor(paste(Temperature, Days, sep = " - ")))
```


Subset to only be push cap, because the screw cap storage was not great. :(

```{r}
dat_push <- dat %>% 
  dplyr::filter(tube == "push")

summary(dat_push)
```









# Statistics


## Marginal Means

Get the average osmolality for each time point/ temperature, and the pairwise statistical differences. 

check the lm assumptions of the 



### lm

Use this lm to determine whether the effects of temperature, time, or their interaction are statistically significant.

```{r}
lm_interaction <- lm(data = dat_push, osmolality_mmol_kg_mean ~ Temperature*Days)

# is the interaction helpful?
lm_single <- lm(data = dat_push, osmolality_mmol_kg_mean ~ Temperature + Days)

# compare
anova(lm_interaction, lm_single)
anova(lm_interaction)
anova(lm_single)
```

The temperature that plasma is stored at and the number of days it is stored *does* affect osmolality values, but the interaction is not significant.



### data wrangling

```{r}
remove <- dat_push %>% 
  dplyr::filter(Days != "0")
to_fix <- dat_push %>% 
  dplyr::filter(Days == "0")
fridge <- to_fix %>% 
  mutate(Temperature = "Refridgerator")
freezer <- to_fix %>% 
  mutate(Temperature = "Freezer")
together <- remove %>% 
  rbind(fridge) %>% 
  rbind(freezer)
```




### lm

Now, run a different lm version, with temperature and days as a single variable, to make the pairwise comparison easier. **I could not figure out how to run this otherwise**

```{r}
lm_arranged <- lm(data = together, osmolality_mmol_kg_mean ~ Temperature*Days)
```




### emmeans

Now, get the marginal means and pairwise differences between each sample group:

```{r}
data.frame(emmeans(lm_arranged, pairwise ~ Temperature | Days)$emmean)

temp_diffs <- data.frame(emmeans(lm_arranged, pairwise ~ Temperature | Days)$contrasts)
temp_diffs

time_diffs <- data.frame(emmeans(lm_arranged, pairwise ~ Days | Temperature)$contrasts) %>% 
  dplyr::filter(substr(contrast, 1, 1) == "0")
time_diffs
```


SE **does** typically vary... interesting...


## Variation Among Samples of the Same Group

Quantify the difference in spread for samples of the same "treatment" group.

```{r}

```




## Change

**I actually think this will be redundant :(**
What was the change for each time point/ temperature, and was it statistically different from zero?

Can I put t-test results into the same df as the change values I calculate?!






# Figure


## Pretty Stuff

```{r}
theme_set(theme_classic())
```



## Means 


```{r}
ggplot() +
  geom_point(data = dat_push,
             aes(x = Days,
                 y = osmolality_mmol_kg_mean,
                 color = Temperature),
             position = position_dodge(width = 0.5))
```








## Change 















